Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-09-15 Manuscript_Nater_etal.qmd
Further text edits

The model of @nilsen2023 integrated line transect data with radio-telemetry data from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored at least once a month by radio triangulation. Most of the fieldwork was conducted from the ground, but to avoid data gaps, the birds were also triangulated from helicopters in May, September, and November. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained fewer observations due to challenging field work conditions. A proportion of the birds were harvested annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switches, we were also able to locate and retrieve a high proportion of birds that died from natural causes, resulting in a known-fate mark-recapture dataset. The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.

In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.

@@

![Graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow.](Figures/Workflow_layout.png){#fig-workflow width="424"}

@@

VRcorrs <- readRDS("Results/VR_corrCoef.rds")
samCorr_mean <- round(subset(VRcorrs, area == "Overall")[,2:3], digits = 3)
```
Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area). In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam). Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").

>>>>

The model of @nilsen2023 integrated line transect data with radio-telemetry data from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. This included males and females, and young-of-the year (8-9 months at capture) and adult (\>1 year old) birds. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored at least once a month by radio triangulation. Most of the fieldwork was conducted from the ground, but to avoid data gaps, the birds were also triangulated from helicopters in May, September, and November. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained fewer observations due to challenging field work conditions. A proportion of the birds were harvested annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switches, we were also able to locate and retrieve a high proportion of birds that died from natural causes, resulting in a known-fate mark-recapture dataset. Several previous studies found no effect of the telemetry devices on ptarmigan survival and further details on the radio-telemetry study can be found in @Israelsen2020 and @Arnekleiv2022.

In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period. We pooled data for males and females as survival was previously found to be very similar [@Israelsen2020] and did not distinguish age classes for analysis.

@@

![Simplified graphical representation of the "targets pipeline" for the multi-area modelling setup. Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow. For the full representation, download the repository and call "targets::tar_visnetwork()". The manual R and Nix/GNUparallel implementations of the workflow have the same structure and built on the same functions and relationships. ](Figures/Workflow_layout.png){#fig-workflow width="424"}

@@

VRcorrs <- readRDS("Results/VR_corrCoef.rds")
samCorr_mean <- round(subset(VRcorrs, area == "Overall")[,2:3], digits = 3)
```

Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area). In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam). Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").
